Typed Clojure

### 2013-07-29 schemeworkshop13.tex
more stuff

\usepackage{mmm}
\usepackage{mathpartir}
\usepackage{clj-grammar}
%\usepackage{url}
%\usepackage[style=alphabetic, sorting=nyt]{biblatex}
%\addbibresource{bibliography.bib}

@@

@@

\title{Typed Clojure}
\subtitle{A Practical Type System for Clojure based on Typed Racket}

\authorinfo{Ambrose Bonnaire-Sergeant}
           {}
           {abonnairesergeant@gmail.com}
\authorinfo{Rowan Davies}
           {University of Western Australia}
           {rowan@csse.uwa.edu.au}

@@

% Things todo
% - small calculus based on previous TR formalisation
%   that demos the ``flow'' environment & not/intersection type
% - discussion and examples for flow filters
% - discussion on the not/difference type, real code where it helps
% - higher-rank polymorphism, show what it enables (eg. monads)
%   - also show more complicated example, making an extensible
%     type for clojure.core/conj, and why it's not straightforward
% - discussion on limitations of local type inference
% - demonstration of how we check multimethods
% - dealing with array covariance
% - note on how we avoid null-pointer exceptions
% - discussion on why Typed Clojure only uses the ``static'' part
%   of Typed Racket
% - the equality filter problem (and draft solution?), a proposition
%   that knows about binding aliasing. Sounds straightforward?

@@

@@

\begin{figure*}
\begin{mathpar}
\infer [T-Var]
{ \inpropenv {\propenv{}} {\isprop {\x{}} {\t{}}}}
{ \judgement {\propenv{}} {\hastype {\x{}} {\t{}}} {\filterset {{\notprop {\falsy} {\x{}}}} {{\isprop {\falsy} {\x{}}}}}
             {\emptyobject{}} {\topprop{}}}

@@

Supporting statically sound interactions with Java arrays is a goal
of Typed Clojure. This is complicated by Java's decision to make
arrays covariant in their argument, a well documented source of static
unsoundness.

Our basic approach is to make our array types \emph{bivariant}. Array types
look like \lstinline|(Array2 w r)| and
are reminiscent of functions or pipes: having a contravariant parameter for input (writing)
and a covariant parameter for output (reading).

Most commonly, an array type is invariant in its parameter; it can
write and read input of the same type.
We can get the same effect by setting our input and output
parameters to the same type. For example, \lstinline|(Array2 Number Number)|
(or equivalently, \lstinline|(Array Number)|)
in Typed Clojure is similar to invariant array types of \lstinline|Number| in languages like Scala.

The biggest gain in using a separate input parameter is the ability
to specify \emph{read-only} arrays. Crucially, our type system features an
explicit bottom type \lstinline|Nothing|, allowing a read-only \lstinline|Number| array
to be of type \lstinline|(Array2 Nothing Number)|.

To realise why defining read-only arrays are useful, we need to examine
what makes array covariance unsound in Java.
Array covariance lets us lie about the type of an array so the consumer
of an array cannot tell the actual type of the array when examining a type
signature.

\begin{verbatim}
...
public static Number[] getNumberArray() {
  Number[] n = new Integer[10];
  return n;
}
...
\end{verbatim}

@@

%\bibliographystyle{abbrvnat}
%
%% The bibliography should be embedded for final submission.
%
%\begin{thebibliography}{}
%\softraggedright
%
%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
%P. Q. Smith, and X. Y. Jones. ...reference text...
%
%\end{thebibliography}

>>>>

\usepackage{mmm}
\usepackage{mathpartir}
\usepackage{clj-grammar}
%\usepackage{url}
%\usepackage{biblatex}
\include{bibliography.bib}

@@

\newcommand\smallsection[1]{\bf \emph{#1}}

@@

\title{Typed Clojure}
\subtitle{A Practical Type System for Clojure, based on Typed Racket}

\authorinfo{Ambrose Bonnaire-Sergeant}
           {University of Western Australia}
           {abonnairesergeant@gmail.com}
\authorinfo{Rowan Davies}
           {University of Western Australia}
           {rowan@csse.uwa.edu.au}

@@

% Things todo
% - DONE small calculus based on previous TR formalisation
%   that demos the ``flow'' environment & not/intersection type
% - discussion and examples for flow filters
% - discussion on the not/difference type, real code where it helps
% - higher-rank polymorphism, show what it enables (eg. monads)
%   - also show more complicated example, making an extensible
%     type for clojure.core/conj, and why it's not straightforward
% - discussion on limitations of local type inference
% - demonstration of how we check multimethods
% - dealing with array covariance
% - note on how we avoid null-pointer exceptions
% - discussion on why Typed Clojure only uses the ``static'' part
%   of Typed Racket
% - the equality filter problem (and draft solution?), a proposition
%   that knows about binding aliasing. Sounds straightforward?

@@

\subsection{Based on Typed Racket}

The design, implementation and theory developed for the Typed Racket project
has proven to be a valuable base for a type system for Clojure.

{\smallsection {Occurrence typing}}
Tobin-Hochstadt and Felleisen~\cite{TF08,TF10} developed \emph{occurrence
typing}, which enables Typed Racket to infer better types at branches.
Typed Clojure uses occurrence typing in a similar way to Typed Racket.
We extend and apply occurrence typing in several interesting ways discussed
in Section [?].

{\smallsection {Practical Variable-Arity Polymorphism}}
Functions with non-trivial variable parameters are common in Racket.
For example, Racket provides \emph{map} which takes a function and a
variable number of collections and applies the function simultaneously
to each element of the provided collections, returning a list of results.
Strickland, Tobin-Hochstadt and Felleisen%~\cite{STF09}
developed a practical system that handles advanced variable parameters
which can handle applications of functions like \emph{map}.

Clojure has a similar emphasis on variable-arity functions. In some ways,
Clojure's core library encourages even more complicated variable-parameter schemes.
The \emph{assoc} core function, for example, takes three parameters and
then a quantity of variable parameters that is a multiple of two.
This is beyond what Typed Racket can currently handle.

Functions like \emph{map} are common in Clojure, so we provide an implementation
of variable-arity polymorphism which has similar capabilities as Typed Racket's
implementation.

{\smallsection {Local Type Inference}}

% TODO references
% why is TR such a good base?
% - immutability
% - common lisp ancestry
% differences?
% - Clojure is built on JVM
% - interop with JVM
% - Clojure's idiomatic primitives are different
% - multimethods + protocols
% - less sophisticated macro system
%  - not an issue
%  - implementation difference, AST walking vs syntax walking

@@

\begin{figure*}
\begin{mathpar}
\infer [T-Var]
{ \inpropenv {\propenv{}} {\isprop {\x{}} {\t{}}}}
{ \judgement {\propenv{}} {\hastype {\x{}} {\t{}}} {\filterset {{\notprop {\falsy} {\x{}}}} {{\isprop {\falsy} {\x{}}}}}
             {\x{}} {\topprop{}}}

@@

Supporting statically sound interactions with Java arrays is a goal
of Typed Clojure. This is complicated by Java's decision to make
arrays covariant in their argument, a well documented source of static
unsoundness. Bracha~\cite{Bra98} summarises Java's approach to maintaining
soundness at runtime, which involves all array writes being checked by
runtime assertions.

This approach fits Java's type system, but we can do better in a more powerful
type system like Typed Clojure. Our goal is to catch all type-incorrect array
writes at compile time so the type system can do more to help Clojure programmers
use arrays, especially those being passed from foreign Java code.

Our basic approach is to make our array types \emph{bivariant}. Array types
look like {\ArrayTwo {\t{w}} {\t{r}}} and
are reminiscent of functions or pipes: having a contravariant parameter for input (writing)
and a covariant parameter for output (reading).
This type can write type {\t{w}} and read type {\t{r}}.

Most commonly, an array type is invariant in its parameter; it can
write and read input of the same type.
We can get the same effect by setting our input and output
parameters to the same type. For example, {\ArrayTwo {\Number} {\Number}}
(or equivalently, {\Array {\Number}}
in Typed Clojure is similar to invariant array types of \Number in languages like Scala.

The biggest gain in using a separate input parameter is the ability
to specify \emph{read-only} arrays. Crucially, our type system features an
explicit bottom type \lstinline|Nothing|, enabling a read-only \lstinline|Number| array
to be of type \lstinline|(Array2 Nothing Number)|.

To realise why defining read-only arrays are useful, we need to examine
what makes array covariance unsound in Java.
Array covariance about the type of an array so the consumer
of an array cannot tell the actual type of the array when examining a type
signature.

\begin{lstlisting}
...
public static Number[] getNumberArray() {
  Number[] n = new Integer[10];
  return n;
}
...
\end{lstlisting}

@@

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\bibliography{bibliography}

%\begin{thebibliography}{}
%\softraggedright
%
%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
%P. Q. Smith, and X. Y. Jones. ...reference text...
%
%\end{thebibliography}
